---
title: "Reproducible Research - Week 2 Assignment"
author: "P Jewell"
date: "19 January 2018"
output: html_document
---


##Loading and preprocessing the data 

1. Load the data
```{r read in dataset}
data <- read.csv("activity.csv")
```
2. Process the data into a suitable format
```{r}
library(lubridate)
data$date <- ymd(data$date)
```


##What is the mean number of steps taken per day?

Total number of steps taken per day
```{r}
Tsteps <- aggregate(data$steps, by = list(Date = data$date), FUN = sum)
names(Tsteps) <- c("Date", "Total")
Tsteps
```

Histogram of total number of steps taken each day
```{r}
library(ggplot2)
hist <- ggplot(data=Tsteps, aes(Total)) +  geom_histogram(binwidth=1000) + xlab("Total steps each day")+ ggtitle ("Histogram of total number of steps taken each day")
hist 
```

Mean and median number of steps taken each day
``` {r}
mean(Tsteps$Total, na.rm=T)
median(Tsteps$Total, na.rm=T)
```


##What is the average daily activity pattern?

Time series plot of the average number of steps taken
```{r}
minterval <- aggregate(data$steps, by=list(interval = data$interval), FUN = mean, na.rm=T)
names(minterval) <- c("interval", "mean.steps")

ggplot(data = minterval,aes(x=interval,y=mean.steps)) + geom_line() + ggtitle("Time series of mean steps taken per 5 minute interval")

```

5 minute interval that, on average, contains the maximum number of steps 
```{r}
minterval[which.max(minterval$mean.steps),]
```


##Imputing missing values

Total number of missing values in the dataset
```{r}
sum(is.na(data$steps))
```

Filling in the missing values in the data set and creating an new data set with missing values filled in 
```{r}
newdata <- data
mean <- tapply(newdata$steps, newdata$interval,mean,na.rm=T)
newdata$steps[is.na(newdata$steps)]  <- mean[as.character(newdata$interval[is.na(newdata$steps)])]

head(newdata)

sum(is.na(newdata))

```


Histogram of total number of steps taken each day after missing values are imputed
```{r}
Tsteps2 <- aggregate(newdata$steps, by = list(Date = newdata$date), FUN = sum)
names(Tsteps2) <- c("Date", "Total")
library(ggplot2)
hist <- ggplot(data=Tsteps2, aes(Total)) +  geom_histogram(binwidth=1000) + xlab("Total steps each day")+ ggtitle ("Histogram of total number of steps taken each day (with missing values replaced)")
hist 

```

Mean and median total number of steps taken per day 
```{r}
mean(Tsteps2$Total)
median(Tsteps2$Total)
```

Difference between these values and originals 
```{r}
mean(Tsteps2$Total)
median(Tsteps2$Total)
mean(Tsteps$Total, na.rm=T)
median(Tsteps$Total, na.rm=T)
```

Impact of imputing missing data
- The impact is that now, with the new data, the mean and medians are the same 


##Differences in activity patterns between weekday and weekends? 

Create a new factor variasble in the dataset with 2 levels - "weekday" and "weekend"
```{r}
library(dplyr)
newdata <- mutate(newdata, day = 
    ifelse(weekdays(newdata$date)=="Saturday"| weekdays(newdata$date)=="Sunday","weekend", "weekday")) 
```

Time series comparing weekends to weekdays
```{r}

newint <- newdata %>%
    group_by(interval,day) %>%
    summarise(steps=mean(steps))

plot <- ggplot(newint,aes(interval,steps)) + geom_line() + facet_wrap(~day)
plot
```

This shows more activity during weekdays than weekends   

END